Quantum-Classical Phase Transition for Exciton-Polariton Josephson Junctions

### 2017-01-19 writing/phase_transition/phase_transition.tex
Tunneling thought the small barrier done.

Inserting (\ref{eq:transformation_of_phi}) into (\ref{eq:schrodinger}) we come to convenient Schr\"odinger equation
%
\begin{equation}
\frac{d^2\Psi}{dz^2} + (E - V(z))\Psi = 0
\label{eq:schrodinger_usual}
\end{equation}
%
for quantum particle with mass $m_{eff} = \dfrac{1}{2}$, normalized energy $E \equiv \dfrac{E}{\alpha}$ moving in the potential $V(z) = s^2 V_0(z)$ with
%
\begin{equation}
V_0(z) = \frac{ (\frac{1}{4} \Lambda^2 - \lambda (1 - \lambda)) \sn^2{z} - \Lambda \cn{z}}{\dn^2{z}}
\label{eq:potential}
\end{equation}
%

@@

{\red
Using the fact that $\lambda$ is small we can replace the potential (\ref{eq:fock1_potential}) with
%
\begin{equation}
V_F(z) \sim -\lambda (1 - \lambda) \sin^2{z},
\end{equation}
%
and our Schrodinger equation transforms to the so-called Mathieu equation
%
\begin{equation}
\frac{d^2 \Psi}{d z^2} + (a - 2q  \cos{2z}) \Psi = 0.
\end{equation}
%
where $a = E + s^2 \lambda(1 - \lambda) / 2$, $q = s^2 \lambda (1 - \lambda)/ 4$.
We recall our boundary condition: $\Psi(z + 2\pi) = \Psi(z)$.
So, spectrum consist of all Mathieu functions $ce_0(z, q)$, $ce_1(z, q)$, $se_1(z, q)$, \dots, which are $2\pi$-periodic, and corresponding eigenvalues $a$ can be found using series expansions in powers of $q$ (see Abramowitz and Stegun).
}

@@

Here we suppose that {\red $\Lambda < 1$}.
This regime represents intermediate case between Rabi and Fock limits --- see Fig. \ref{phase_potential}b.
In general all terms in (\ref{eq:potential}) should be kept.
The behavior of effective potential $V_0(z)$ depends on the ratio between parameters $\Lambda$ and $\lambda$.
If $\Lambda < 2\lambda$ the $V_0(z)$ posses minima at $z = 0$ and $V(0) = -\Lambda$.
Frequency of small amplitude ($z \ll 1$) oscillations can be find from (\ref{eq:potential}) and looks like
{\red equation for small oscillations}

In the other limit, for $2 \lambda > \Lambda$, two minima of $V_0(z)$ with coordinates
%
\begin{equation}
\cn{z_{min}} = \frac{\Lambda (2 \lambda - 1)}{0.5 \Lambda^2 - 2 \lambda (1 - \lambda)}
\label{eq:z_minima}
\end{equation}
%
appears both for Josephson and Fock regimes, cf. Fig. \ref{phase_potential}b,c.  
The depth of the potential minima depends on the $\lambda$ parameter.
It is interesting to note that for {\red $\Lambda > 1 + 2\lambda + \sqrt{1 + 8\lambda}$} potential $V_0(z)$ presumes local minimum at {\red $\pm 2 K(\lambda)$}.

Practically two minima located at the points $A$ and $B$ of $W$-shape quantum phase potential in Fig. \ref{phase_potential}b can be used for initialization of exciton polariton qubit states $\ket{0}$ and $\ket{1}$ similarly to superconductor flux qubits, respectively, {\red cf.[ ]}.
However, in our case such qubits much more easier to tailor by external optical or electrical pump {\red [ ]}.
Another interesting phenomena that can be obtained with potential in Fig. \ref{phase_potential}b is connected with some analogies in cosmology and quantum field theory.
In particular, in our recent work {\red [ ]} we discuss Higgs field generation by using exciton polaritons in semiconductor Bragg mirror structures as an on table analog of known cosmological phenomenon, {\red cf. [ ]}.
In this paper we continue our physical analogies with false vacuum decay (marked by the point $C$ in Fig. \ref{phase_potential}b) that can be established here, {\red cf. [ ]}.

@@

We determiine the temperature of the transition like Owerre does.
At very low temperature, transition occur by quantum tunneling through the barrieer and the rate is governed by $\Gamma \sim e^{-B}$, where $B$ is the instanton action.
At high temperature the particle has the possibility of hopping over the barrier, in this case the rate is governed by $\Gamma \sim e^{-\Delta V / T}$, where $\Delta V$ is the energy barrier.
At a critical point these two transition rate are equal, there exist a crossover (first-order transition) temperature $T_{I}$ from quantum to a thermal regime, it's estimated as $T_{I} = \Delta V / B$.

The crossover temperature in the case of second-order transition can be estimated as $T_{II} = \omega_0 / 2 \pi$, where $\omega_0$ is a oscillation frequency of the thermon near the top of the barrier.

First, let's find out the dependence of the order on the parameters $\lambda$, $\Lambda$.
Expand the potential $V_0(z)$ into a Taylor series:
%
\begin{equation}
\begin{array}{lcl}
V_0(z) & = & \left( \frac{1}{4}\Lambda^2 - \lambda (1 - \lambda) \right) \left( z^2 + \frac{2\lambda - 1}{3} z^4 + o(z^4) \right) \\ && - \Lambda \left( 1 + \frac{2\lambda - 1}{2} z^2 + \frac{16\lambda^2 - 16\lambda + 1}{24} z^4 + o(z^4) \right) \\
& = & -\Lambda + \left( \frac{1}{4} \Lambda^2  - \frac{2\lambda - 1}{2} \Lambda - \lambda (1 - \lambda) \right) z^2 \\
&& + \left( \frac{2\lambda - 1}{12} \Lambda^2 - \frac{16\lambda^2 - 16\lambda + 1}{24} \Lambda - \frac{\lambda (\lambda - 1) (2\lambda - 1)}{3} \right) z^4 + o(z^4).
\end{array}
\label{eq:potential_teylor}
\end{equation}
%
Classical equation of motion for the thermon reads as
%
\begin{equation}
\ddot{z} = s^2 \frac{d V_0}{dz}.
\label{eq:thermon}
\end{equation}
%
Equation for the small ocsillation near the top of the barrier:
%
\begin{equation}
\ddot{z} - s^2 \left( \tfrac{1}{2} \Lambda^2 - (2\lambda - 1) \Lambda - 2 \lambda (1 - \lambda) \right) z = 0.
\label{eq:thermon_small}
\end{equation}
%
It follows from (\ref{eq:thermon_small}) that, as mentioned above, when $\Lambda \ge 2\lambda$ there is no barrier at $z = 0$.

@@

Consider the imaginary-time action
%
\begin{equation}
S(T) = \oint (\tfrac{1}{2} \dot{z}^2 + s^2 V_0(z)) d \tau
\label{eq:thermon_action_1}
\end{equation}
%
Periodic, with $\tau_p = \hbar / k_B T$, solutions of (\ref{eq:thermon}) satisfy the relation
%
\begin{equation}
\tfrac{1}{2} \dot{z}^2 = s^2 V_0(z) - E(\tau_p),
\label{eq:total_energy}
\end{equation}
%
where $\tau_p$ is a thermon period corresponding to the energy level $E(\tau_p)$;
%
\begin{equation}
\tau_p(E) = \sqrt{2} \int_{z_1(E)}^{z_2(E)} \sqrt{s^2 V_0(z) - E} ~dz,
\label{eq:thermon_period}
\end{equation}
%
here $z_{1,2}(E)$ -- turning points.
Using (\ref{eq:total_energy}) rewrite (\ref{eq:thermon_action_1}) in a form 
%
\begin{equation}
S(T) = 2\sqrt{2} \int_{z_1(E)}^{z_2(E)} \sqrt{s^2 V_0(z) - E} ~dz + E \tau_p (E).
\label{eq:thermon_action_2}
\end{equation}
%
We draw dependence $S(T)$ and $\tau_p(E)$ for the cases of the 1st and 2nd order phase transition.
For the sake of convenience we suppose that $V_0(z) = 0$ at minima surrounding the barrier.
So, $E = 0$ corrensponds to the separatrices instanton solution (May be we should make corresponding transformation earlier).
%
\begin{figure}[ht]
\center{\includegraphics[width=\linewidth]{pic/action_period_combined.eps}}
\caption{(a) 1st-order transition, $\lambda = 0.9$, $\Lambda = 0.1$; (b) 2nd-order transition, $\lambda = 0.5$, $\Lambda = 0.5$}
\end{figure}
%

Now let's compute the temperature of the phase transition.
It follows from (\ref{eq:thermon_small}) that the frequency of small thermon oscillation near the top of the barrier can be computted as follows
%
\begin{equation}
\omega_0 = s \sqrt{2 \lambda (1 - \lambda) + (2 \lambda - 1) \Lambda - \tfrac{1}{2} \Lambda^2}.
\label{eq:period_small}
\end{equation}
%
The temperature of the 2nd-order transition can be estimated as
\begin{equation}
T_{II} = (s  / 2 \pi) \sqrt{2 \lambda (1 - \lambda) + (2 \lambda - 1) \Lambda - \tfrac{1}{2} \Lambda^2}.
\end{equation}

>>>>

Inserting (\ref{eq:transformation_of_phi}) into (\ref{eq:schrodinger}) we come to convenient Schr\"odinger equation
%
\begin{equation}
\frac{1}{2 m} \frac{d^2\Psi}{dz^2} + (E - V(z))\Psi = 0
\label{eq:schrodinger_usual}
\end{equation}
%
for quantum particle with ``mass'' $m = \dfrac{1}{2 \alpha}$, energy $E$, moving in the potential $V(z) = \alpha s^2 V_0(z)$ with
%
\begin{equation}
V_0(z) = \frac{ (\frac{1}{4} \Lambda^2 - \lambda (1 - \lambda)) \sn^2{z} - \Lambda \cn{z}}{\dn^2{z}}
\label{eq:potential}
\end{equation}
%

@@

{\red
Using the fact that $\lambda$ is small we can replace the potential (\ref{eq:fock1_potential}) with
%
\begin{equation}
V_F(z) \sim -\lambda (1 - \lambda) \sin^2{z},
\end{equation}
%
and our Schrodinger equation transforms to the so-called Mathieu equation
%
\begin{equation}
\frac{d^2 \Psi}{d z^2} + (a - 2q  \cos{2z}) \Psi = 0.
\end{equation}
%
where $a = E / \alpha + s^2 \lambda(1 - \lambda) / 2$, $q = s^2 \lambda (1 - \lambda)/ 4$.
We recall our boundary condition: $\Psi(z + 2\pi) = \Psi(z)$.
So, spectrum consist of all Mathieu functions $ce_0(z, q)$, $ce_1(z, q)$, $se_1(z, q)$, \dots, which are $2\pi$-periodic, and corresponding eigenvalues $a$ can be found using series expansions in powers of $q$ (see Abramowitz and Stegun).
}

@@

Here we suppose that {\red $\Lambda < 1$}.
This regime represents intermediate case between Rabi and Fock limits --- see Fig. \ref{phase_potential}b.
In general all terms in (\ref{eq:potential}) should be kept.
The behavior of effective potential $V_0(z)$ depends on the ratio between parameters $\Lambda$ and $\lambda$.
If $\Lambda \ge 2\lambda$ the $V_0(z)$ posses minima at $z = 0$ and $V(0) = -\Lambda$.
Frequency of small amplitude ($z \ll 1$) oscillations can be find from (\ref{eq:potential}) and looks like {\red equation for small oscillations}.

In the other limit, for $\Lambda < 2 \lambda$, two minima of $V_0(z)$ with coordinates
%
\begin{equation}
\cn{z_{min}} = \frac{\Lambda}{2 \lambda}
\label{eq:z_minima}
\end{equation}
%
appears both for Josephson and Fock regimes, cf. Fig. \ref{phase_potential}b,c.  
The depth of the potential minima depends on the $\lambda$ parameter.
It is interesting to note that for {\red $\Lambda > 1 + 2\lambda + \sqrt{1 + 8\lambda}$} potential $V_0(z)$ presumes local minimum at $\pm 2 K(\lambda)$.

Practically two minima located at the points $A$ and $B$ of $W$-shape quantum phase potential in Fig. \ref{phase_potential}b can be used for initialization of exciton polariton qubit states $\ket{0}$ and $\ket{1}$ similarly to superconductor flux qubits, respectively, {\red cf. [ ]}.
However, in our case such qubits much more easier to tailor by external optical or electrical pump {\red [ ]}.
Another interesting phenomena that can be obtained with potential in Fig. \ref{phase_potential}b is connected with some analogies in cosmology and quantum field theory.
In particular, in our recent work {\red [ ]} we discuss Higgs field generation by using exciton polaritons in semiconductor Bragg mirror structures as an on table analog of known cosmological phenomenon, {\red cf. [ ]}.
In this paper we continue our physical analogies with false vacuum decay (marked by the point $C$ in Fig. \ref{phase_potential}b) that can be established here, {\red cf. [ ]}.

@@

We determiine the temperature of the transition like Owerre does.
At very low temperature, transition occur by quantum tunneling through the barrieer and the rate is governed by $\Gamma \sim e^{-B}$, where $B$ is the instanton action.
At high temperature the particle has the possibility of hopping over the barrier, in this case the rate is governed by $\Gamma \sim e^{-\Delta V / T}$, where $\Delta V$ is the energy barrier.
At a critical point these two transition rate are equal, there exist a crossover (first-order transition) temperature $T_{c}^{(1)}$ from quantum to a thermal regime, it's estimated as $T_{c}^{(1)} = \Delta V / B$.

The crossover temperature in the case of second-order transition can be estimated as $T_{c}^{(2)} = \omega_0 / 2 \pi$, where $\omega_0$ is a oscillation frequency of the thermon near the top of the barrier.

First, let's find out the dependence of the order on the parameters $\lambda$, $\Lambda$.
Expand the potential $V_0(z)$ into a Taylor series:
%
% \begin{equation}
% \begin{array}{lcl}
% V_0(z) & = & \left( \frac{1}{4}\Lambda^2 - \lambda (1 - \lambda) \right) \left( z^2 + \frac{2\lambda - 1}{3} z^4 + o(z^4) \right) \\ && - \Lambda \left( 1 + \frac{2\lambda - 1}{2} z^2 + \frac{16\lambda^2 - 16\lambda + 1}{24} z^4 + o(z^4) \right) \\
% & = & -\Lambda + \left( \frac{1}{4} \Lambda^2  - \frac{2\lambda - 1}{2} \Lambda - \lambda (1 - \lambda) \right) z^2 \\
% && + \left( \frac{2\lambda - 1}{12} \Lambda^2 - \frac{16\lambda^2 - 16\lambda + 1}{24} \Lambda - \frac{\lambda (\lambda - 1) (2\lambda - 1)}{3} \right) z^4 + o(z^4).
% \end{array}
% \label{eq:potential_teylor}
% \end{equation}
%
\begin{equation}
\begin{array}{lcl}
V_0(z) & = & -\Lambda + \left( \frac{1}{4} \Lambda^2  - \frac{2\lambda - 1}{2} \Lambda - \lambda (1 - \lambda) \right) z^2 \\
&& + \left( \frac{2\lambda - 1}{12} \Lambda^2 - \frac{16\lambda^2 - 16\lambda + 1}{24} \Lambda - \frac{\lambda (\lambda - 1) (2\lambda - 1)}{3} \right) z^4 + o(z^4).
\end{array}
\label{eq:potential_teylor}
\end{equation}
%
Classical equation of motion for the thermon reads as
%
\begin{equation}
m \ddot{z} = \alpha s^2 \frac{d V_0}{dz}.
\label{eq:thermon}
\end{equation}
%
Equation for the small ocsillation near the top of the barrier:
%
\begin{equation}
m \ddot{z} - \alpha s^2 \left( \tfrac{1}{2} \Lambda^2 - (2\lambda - 1) \Lambda - 2 \lambda (1 - \lambda) \right) z = 0.
\label{eq:thermon_small}
\end{equation}
%
It follows from (\ref{eq:thermon_small}) that, as mentioned above, when $\Lambda \ge 2\lambda$ there is no barrier at $z = 0$.

@@

Consider the imaginary-time action
%
\begin{equation}
S(T) = \oint (\tfrac{1}{2} m \dot{z}^2 + V(z)) d \tau
\label{eq:thermon_action_1}
\end{equation}
%
Periodic, with $\tau_p = \hbar / k_B T$, solutions of (\ref{eq:thermon}) satisfy the relation
%
\begin{equation}
\tfrac{1}{2} m \dot{z}^2 = V(z) - E(\tau_p),
\label{eq:total_energy}
\end{equation}
%
where $\tau_p$ is a thermon period corresponding to the energy level $E(\tau_p)$;
%
\begin{equation}
\tau_p(E) = \sqrt{2 m} \int_{z_1(E)}^{z_2(E)} \sqrt{V(z) - E} ~dz,
\label{eq:thermon_period}
\end{equation}
%
here $z_{1,2}(E)$ -- turning points.
Using (\ref{eq:total_energy}) rewrite (\ref{eq:thermon_action_1}) in a form 
%
\begin{equation}
S(T) = 2 \sqrt{2 m} \int_{z_1(E)}^{z_2(E)} \sqrt{V(z) - E} ~dz + E \tau_p (E).
\label{eq:thermon_action_2}
\end{equation}
%
We draw dependence $S(T)$ and $\tau_p(E)$ for the cases of the 1st and 2nd order phase transition.
For the sake of convenience we suppose that $V_0(z) = 0$ at minima surrounding the barrier.
So, $E = 0$ corrensponds to the separatrices instanton solution (May be we should make corresponding transformation earlier).
%
\begin{figure}[ht]
\begin{minipage}[h]{0.49\linewidth}
\center{\includegraphics[width=1\linewidth]{pic/action_period_1st_order_transition.eps} \\ (a)}
\end{minipage}
\hfill
\begin{minipage}[h]{0.49\linewidth}
\center{\includegraphics[width=1\linewidth]{pic/action_period_2nd_order_transition.eps} \\ (b)}
\end{minipage}
\caption{(a) 1st-order transition, $\lambda = 0.9$, $\Lambda = 0.1$; (b) 2nd-order transition, $\lambda = 0.5$, $\Lambda = 0.5$; parameters: $\hbar = 1$, $k_B = 1$, $\alpha = 1$, $s = 1$; the solid liine correspond to the thermon action, $S_T(T)$; the dashed line corresponds to the thermodynamic action, $S_0(T) = \hbar \Delta V / k_b T$. \label{pic:action_period}}
\end{figure}
%

Now let's compute the temperature of the phase transition.
It follows from (\ref{eq:thermon_small}) that the frequency of small thermon oscillation near the top of the barrier can be computted as follows
%
\begin{equation}
\omega_0 = 2 \alpha s \sqrt{\lambda (1 - \lambda) + \tfrac{1}{2} (2 \lambda - 1) \Lambda - \tfrac{1}{4} \Lambda^2}.
\label{eq:period_small}
\end{equation}
%
The temperature of the 2nd-order transition can be computed as
%
\begin{equation}
T_{c}^{(2)} = (\alpha s / \pi) \sqrt{\lambda (1 - \lambda) + (2 \lambda - 1) \tfrac{\Lambda}{2} - \tfrac{1}{4} \Lambda^2}.
\end{equation}
%

The temperature of the 1st-order transition can be estimated as
%
\begin{equation}
T_{c}^{(1)} = \frac{\Delta V}{B},
\end{equation}
%
where $\Delta V$ -- high of the potential barrier and $B$ is a instanton action which can be calculated as $B = S(E_{min})$.
Excluding some constant form the hamiltonian we can rewrite our potential in a form
%
\begin{equation}
\widetilde{V}_0(z) = \frac{\lambda (\cn{z} - \tfrac{\Lambda}{2 \lambda})^2}{\dn^2{z}}.
\end{equation}
%
It's convenient, because when the barrier exist at the point $z = 0$, ($\Lambda < 2 \lambda$), $\widetilde{V}_0(z_{min}) = 0$ and $E_{min} = 0$.
Action of the instanton trajectory can be obtained by integration
%
\begin{equation}
\begin{array}{c}
B = S(E_{min}) = 2 \sqrt{2 m} \int_{-z_{min}}^{+z_{min}} \sqrt{\alpha s^2 \widetilde{V}_0(z)} ~dz = \\
= 2 s \left[ \log \frac{2 \sqrt{\lambda} + \sqrt{4 \lambda^2 - \Lambda^2}}{2 \sqrt{\lambda} - \sqrt{4 \lambda^2 - \Lambda^2}} - \frac{\Lambda}{\sqrt{\lambda (1 - \lambda)}} \arctan \left( \frac{\sqrt{(1 - \lambda) (4 \lambda^2 - \Lambda^2)}}{\Lambda} \right) \right];
\end{array}
\end{equation}
%
here $z_{min} = \frac{\Lambda}{2 \lambda}$.
So, $T_{c}^{(1)} = \frac{\Delta V}{B}$, $\Delta V = \alpha s^2 \widetilde{V}_0(0) = \alpha s^2 \lambda (1 - \frac{\Lambda}{2 \lambda})^2$.
You can see on Fig.\ref{pic:action_period}(a) an example of such estimation (left red line).
It's differ form real value of tempereture (right red line) by $\sim 1.2\%$.
